---
title: "Assignment 3: Verify the Claim"
subtitle: "Diving into the census data to verify the claims of increased white poverty in Philadelphia"
author: "Isaac Tan"
date: 2026-09-25
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

## The Question

In 2018, WHYY published an article [New Census figures on Philly
neighborhoods show inequality, high numbers of whites living in
poverty](https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/),
stating many claims about the changes to the state of poverty and
inequality in the city. While the claims in the article are supported
with data from the ACS5, the author only briefly mentions the margins of
error embedded within the data that has the potential to undermine their
findings. In this report, I will dive into the margins of error between
the two data sets to determine if the conclusions drawn within the
article are statistically significant.

------------------------------------------------------------------------

## Part 1: Reconstructing the Data

We begin first with sourcing the ACS data through tidycensus.


```{r}
#| include: false
library(tidyverse)
library(tidycensus)
library(sf)
library(ggplot2)
library(gt)
```

```{r}
#| include: false
state <- "PA"
county <- "Philadelphia"
var_income <- "B19013_001"
var_below_poverty <- "B17001_002"
var_total_poverty_summary <- "B17001_001"


phl_data_2010 <- get_acs(
  geography = "tract",
  variables = c(income = var_income,
                poverty = var_below_poverty,
                poverty_summary = var_total_poverty_summary),
  state = state,
  county = county,
  year = 2010,
  survey = "acs5",
  geometry = TRUE,
  output = "wide"
)

phl_data_2017 <- get_acs(
  geography = "tract",
  variables = c(income = var_income,
                poverty = var_below_poverty,
                poverty_summary = var_total_poverty_summary),
  state = state,
  county = county,
  year = 2017,
  survey = "acs5",
  geometry = TRUE,
  output = "wide"
)

```

### Variables

The variables selected in the reconstruction of their analysis are:

| Variable | ACS5 Code |
|------------------------------------|------------------------------------|
| Median Household Income in the past 12 months | B19013_001 |
| Poverty Status (Income in the past 12 months below poverty level) | B17001_002 |
| Summary of Poverty Status Population | B17001_001 |

### ACS Selection

The ACS5 was selected for this project to be in line with the analysis
from the article. Similarly, the two periods selected were 2006-2010 and
2013-2017. As the ACS5 data is aggregated across a period of 5 years,
the singular selection of ACS5 2010 and ACS5 2017 maps onto the time
periods stated above.

### Assumptions

The assumptions of this project are that the census boundaries and
methods did not change between the two time periods chosen. It was also
assumed that the census tract boundaries and the neighborhoods used for
comparison were comparable. This assumption had to be made because the
article's scope of analysis utilized unofficial neighborhood maps due to
their easily identifiable nature by Philadelphia residents.

It was noted that the two periods in the article (2006-2010 and
2013-2017) do not overlap. This is crucial to ensure that the data being
compared is not counted twice within the analysis.

## Part 2: Testing the Data

```{r}
#| results: hide
phl_data_2010 <- phl_data_2010 %>%
  mutate(
    poverty_pct = povertyE / poverty_summaryE * 100,
    poverty_moe_pct = moe_prop(
      num       = povertyE,
      denom     = poverty_summaryE,
      moe_num   = povertyM,
      moe_denom = poverty_summaryM
    ) * 100
  )

phl_data_2017 <- phl_data_2017 %>%
  mutate(
    poverty_pct = povertyE / poverty_summaryE * 100,
    poverty_moe_pct = moe_prop(
      num       = povertyE,
      denom     = poverty_summaryE,
      moe_num   = povertyM,
      moe_denom = poverty_summaryM
    ) * 100
  )
```

### ACS Data for 2006-2010
```{r}
head(phl_data_2010)
```

### ACS Data for 2013-2017

```{r}
head(phl_data_2017)
```

For each census tract in both periods, the poverty rate was computed as
the share of the tract population falling below the federal poverty
line. The margin of error for this rate was derived using `moe_prop()`.

When opening the data, it is observed that certain census tracts return
a `NaN` value for the chosen variables. When matched with the census
tract shapefiles and the map of Philadelphia, most of these tracts are
aligned with the Airports (Philadelphia International Airport and the
Northeast Philadelphia Airport) and parks that do not have high
residential populations, thus resulting in a lack of data. These were
left in the dataset to ensure the geometry of the Philadelphia area
remained recognizable to the general public.

### Calculating the Coefficient of Variation and the Reliability

```{r}

# CV and reliability

phl_data_2010 <- phl_data_2010 %>%
  mutate(
    cv = (poverty_moe_pct / 1.645) / poverty_pct * 100,
    reliability = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat Reliable",
      TRUE     ~ "Unreliable"
    )
  )

phl_data_2017 <- phl_data_2017 %>%
  mutate(
    cv = (poverty_moe_pct / 1.645) / poverty_pct * 100,
    reliability = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat Reliable",
      TRUE     ~ "Unreliable"
    )
  )

```

To assess how much weight can be placed on each tract's estimate, a
coefficient of variation (CV) was computed by converting the MOE back to
a standard error and expressing it as a percentage of the estimate.
Tracts were then flagged into three reliability tiers following
Jurjevich et al. (2018): CVs below 12% are considered reliable, between
12% and 40% somewhat reliable, and above 40% unreliable.

The tables below summarize tract-level reliability for both periods. A
large share of tracts fall into the somewhat reliable or unreliable
categories, particularly in low-population areas, which is a known
limitation of ACS5 data due to sampling limitations and varied
populations across census tracts.


#### For 2006-2010

```{r}

# Creating the reliability tables

library(gt)

reliability_table_2010 <- phl_data_2010 %>%
  st_drop_geometry() %>%
  mutate(
    dot = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    flag = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
    NAME = str_remove(NAME, ", Philadelphia County, Pennsylvania")
  ) %>%
  arrange(flag) %>%
  select(dot, NAME, poverty_pct, poverty_moe_pct, cv, flag) 
  

reliability_table_2010 %>%
  gt() %>%
  fmt_number(columns = c(poverty_pct, poverty_moe_pct), decimals = 0) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_label(
    dot      = "",
    NAME     = "County",
    poverty_pct = "Estimate of Poverty Percent",
    poverty_moe_pct      = "MOE",
    cv       = "CV (%)",
    flag     = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title    = "Population below 100% of the poverty level",
    subtitle = "Philadelphia counties, ACS 2010 5-year estimates (2006-2010)"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )

```

#### For 2013-2017

```{r}
reliability_table_2017 <- phl_data_2017 %>%
  st_drop_geometry() %>%
  mutate(
    dot = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    flag = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
    NAME = str_remove(NAME, ", Philadelphia County, Pennsylvania")
  ) %>%
  arrange(flag) %>%
  select(dot, NAME, poverty_pct, poverty_moe_pct, cv, flag) 
  

reliability_table_2017 %>%
  gt() %>%
  fmt_number(columns = c(poverty_pct, poverty_moe_pct), decimals = 0) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_label(
    dot      = "",
    NAME     = "County",
    poverty_pct = "Estimate of Poverty Percent",
    poverty_moe_pct      = "MOE",
    cv       = "CV (%)",
    flag     = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title    = "Population below 100% of the poverty level",
    subtitle = "Philadelphia counties, ACS 2017 5-year estimates (2013-2017)"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )

```


### Testing the Results for Statistical Significance

To test whether observed changes between the two periods are
statistically meaningful, the two datasets were combined into a single
wide-format table with one row per tract. The change in poverty rate
between 2006–2010 and 2013–2017 was then computed alongside its margin
of error. Following the ACS convention that published MOEs reflect 90%
confidence intervals, each MOE was first converted to a standard error
by dividing by 1.645 (the z-estimate for a 90% confidence interval). The
standard error of the difference was derived by taking the square root
of the sum of squared standard errors from each period (the standard
formula for the uncertainty of a difference between two independent
estimates). A z-score was then calculated by dividing the change by its
standard error. A tract's change is considered statistically
distinguishable from zero at 90% confidence when the absolute z-score
exceeds 1.645. The same procedure was applied to median household income
as well.

```{r}
# Binding results

poverty_wide <- bind_rows(
  phl_data_2010 %>%
    st_drop_geometry() %>%
    select(GEOID, NAME, poverty_pct, poverty_moe_pct) %>%
    mutate(period = "2006-2010"),
  phl_data_2017 %>%
    st_drop_geometry() %>%
    select(GEOID, NAME, poverty_pct, poverty_moe_pct) %>%
    mutate(period = "2013-2017")
) %>%
  pivot_wider(
    names_from  = period,
    values_from = c(poverty_pct, poverty_moe_pct)
  )


# testing results

results_poverty <- poverty_wide %>%
  mutate(
    change      = `poverty_pct_2013-2017` - `poverty_pct_2006-2010`,
    se_1        = `poverty_moe_pct_2006-2010` / 1.645,
    se_2        = `poverty_moe_pct_2013-2017` / 1.645,
    se_diff     = sqrt(se_1^2 + se_2^2),
    moe_diff    = se_diff * 1.645,
    z           = change / se_diff,
    significant = abs(z) > 1.645
  )

results_poverty %>%
  count(significant)

```

Of the 384 tracts, 63 show a statistically detectable change in poverty
rate at 90% confidence, while 313 do not. This means the majority of
tract-level changes reported in the article cannot be distinguished from
noise in the underlying survey data.

```{r}
# Calculating the percentage change in income


income_wide <- bind_rows(
  phl_data_2010 %>%
    st_drop_geometry() %>%
    select(GEOID, NAME, incomeE, incomeM) %>%
    mutate(period = "2006-2010"),
  phl_data_2017 %>%
    st_drop_geometry() %>%
    select(GEOID, NAME, incomeE, incomeM) %>%
    mutate(period = "2013-2017")
) %>%
  pivot_wider(
    names_from  = period,
    values_from = c(incomeE, incomeM)
  )

results_income <- income_wide %>%
  mutate(
    change      = `incomeE_2013-2017` - `incomeE_2006-2010`,
    se_1        = `incomeM_2006-2010` / 1.645,
    se_2        = `incomeM_2013-2017` / 1.645,
    se_diff     = sqrt(se_1^2 + se_2^2),
    moe_diff    = se_diff * 1.645,
    z           = change / se_diff,
    significant = abs(z) > 1.645
  )

results_income %>%
  count(significant)

# add geometry to both

results_poverty <- phl_data_2017 %>%
  select(GEOID, geometry, cv, reliability) %>%
  left_join(results_poverty, by = "GEOID")

results_income <- phl_data_2017 %>%
  select(GEOID, geometry, cv, reliability) %>%
  left_join(results_income, by = "GEOID")

```

Similarly, for income, only 78 out of 384 tracts show a statistically
detectable change in median income at 90% confidence, while 296 do not.

### Visualizing the Data

```{r}
# Visualizing the poverty change
library(sf)
library(ggplot2)

#| fig-height:9
results_poverty %>%
  ggplot() +
    geom_sf(aes(fill = change)) +
    theme_minimal() + 
    theme_void() +
    labs( # adjusts the labels
      title = "Changes in Poverty Rate in Philadelphia \nbetween 2006-2010 and 2013-2017",
      subtitle = "People below 100% of the poverty level, ACS 5-year estimates")
```

```{r}
# Visualizing the income change

#| fig-height:9
results_income %>%
  ggplot() +
    geom_sf(aes(fill = change)) +
    theme_minimal() + 
    theme_void() +
    labs( # adjusts the labels
      title = "Changes in Median Income in Philadelphia \nbetween 2006-2010 and 2013-2017",
      subtitle = "ACS 5-year estimates")

```

```{r}

# Visualize with error bars
results_poverty %>%
  filter(significant) %>%
  arrange(desc(significant)) %>%
  mutate(NAME = str_remove(NAME, ", Philadelphia County, Pennsylvania")) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(NAME, change), y = change)) +
  geom_col(aes(fill = change > 0)) +
  geom_errorbar(aes(ymin = change - moe_diff,
                    ymax = change + moe_diff), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Statistically significant change in poverty count",
       subtitle = "2006–2010 vs 201–2017, ACS 5-year estimates",
       x = NULL, y = "Change in people below poverty level")

```

The error bar chart illustrates the top 15 statistically significant changes in poverty count with their related census tracts.

## Part 3: Adjudicate

### The Article's Claims

The article makes many claims regarding the changes to Philadelphia's
poverty rate, white poverty, and income levels. The section below sorts
out the different claims by three distinctions: 1) Supported by the
Data, 2) Not Supported by the data, and 3) Cannot be Determined

#### 1) Supported by the Data

The article makes a claim that Fairhill is the neighborhood with the
highest inequality based on percentage in poverty (61%) based on the
ACS5 2017 data. Drawing from our own analysis of the data through the
census tracts, the average poverty rate of the three census tracts that
best fit the Fairhill neighborhood boundary is 61.5%, thus demonstrating
that the claims in the article regarding the high inequality in Fairhill
are supported by the data.

```{r}
fairhill <- phl_data_2017 %>%
  filter(GEOID %in% c("42101017500", "42101017601", "42101017602"))

mean_fairhill <-
  mean(fairhill$poverty_pct)

print(mean_fairhill)
```

#### 2) Not Supported by the Data

The article goes in depth into the rise of poverty in the neighborhood
of Eastwick. From their data, the authors claim that there was a rise of
14 percentage points in the poverty rate in Eastwick: Rising from a 12%
poverty rate in 2010 to a 26% poverty rate in 2017. When comparing to
the ACS data at the census tract level used in this project, it is
observed that the average of the four census tracts the most closely
aligns with the neighborhood of Eastwick results in a 17% increase in
the percentage of poverty that is higher than the reported figures in
the article.

```{r}
eastwick <- results_poverty %>%
  filter(GEOID %in% c("42101005400", "42101005600", "42101005500", "42101006000" ))

mean_eastwick <- 
  mean(eastwick$change)

print(mean_eastwick)

```

#### 3) Cannot be Determined

The claims stated by the article regarding white poverty cannot be
determined as there is no race-disaggregated data relating to the
poverty rate. Some of the article's other claims relating to specific
neighborhoods that do not line up with census tract boundaries also
cannot be determined.

## Part 4: Rewrite

New census data confirms that poverty in Philadelphia remains deeply
concentrated, with tracts in the Fairhill area of North Philadelphia
showing poverty rates averaging around 61 percent. This is among the
highest in the city. However, the picture is more uncertain than the
headline figures in the article suggest. Of Philadelphia's 384 census
tracts, only 63 show a change in poverty rate between 2006–2010 and
2013–2017 that is statistically distinguishable from zero at 90%
confidence. For the remaining 313 tracts, the margins of error in the
underlying survey data are wide enough that the apparent changes in
poverty rate cannot be reliably separated from sampling noise. Thus,
claims about dramatic poverty increases in specific neighborhoods, and
about surging white poverty across the Northeast, rest on point
estimates that the data cannot confidently support at the tract level.
