---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Isaac Tan"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

------------------------------------------------------------------------

## The question

Every number the American Community Survey publishes is an estimate
drawn from a sample, not a count. That means every number comes with a
margin of error — a statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

**In this analysis, I look at Median Home Value across counties in
Illinois, and ask which communities we measure most and least
reliably.**

------------------------------------------------------------------------

## Choosing what to measure

```{r}
my_state <- "IL"        # two-letter abbreviation, in quotes
my_variable <- "B25077_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

**The function returned 102 rows, which matches the number of counties
within Illinois**

------------------------------------------------------------------------

## How big is the margin of error?

A margin of error of 3,000 means something very different for an
estimate of 20,000 than for an estimate of 200,000. To compare across
places, we need it as a percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

**Hardin County is the county with the largest margin of error relative
to its estimate at 34.09924.**

------------------------------------------------------------------------

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

**From the calculations, 49 counties fall under 'High Confidence', 42
counties fall under 'Moderate', and 11 counties fall under 'Low
Confidence'.**

------------------------------------------------------------------------

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the
same row. That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability
category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```

**This table illustrates a pattern between average population and the
level of confidence we have in the Median Home Value estimates. The
average population of counties with high confidence was 240,766.12,
while the average population of counties with low confidence was
7350.818. Counties with moderate levels of confidence had an average
population of 19,387. The pattern is generally that with higher
populations, the level of confidence in the data increases.**

------------------------------------------------------------------------

## What this means for policy

1.  *We
    know the least about the 11 counties with the lowest confidence
    levels. This would be Pope County, Bond County, Scott County, Massac
    County, Pulaski County, Calhoun County, Brown County, Hamilton
    County, Jasper County, Alexander County, and Hardin County, which
    had a margin of error percent range of 10.275 to 34.099. As seen in
    the above table, these counties also have lower average populations
    compared to the rest of the state.*
2.  *As the ACS5
    already measures a slice of the county population, smaller counties
    like Hardin County (with about a population of 3623) will have
    smaller samples that lead to higher variability in the median. With
    a small population, the statistical standard error increases,
    resulting in imprecision within the data. Hence, we can conclude
    that the states with the lower populations are the communities that
    we know the least about from the ACS5 data.*
3.  *If a state agency
    allocates funding using the variable of median home values, there is
    a high probability that the ACS5 estimate of the true median of a
    county may be higher than the actual value. This error in the
    estimate may result in low confidence counties appearing wealthier
    than they actually are and receiving less aid than they would need,
    thus resulting in an inequitable provision of welfare funding.*



